# Setting-Up the Software Environment
*Y Yigini*

This cookbook focuses on SOC modeling using open source digital mapping tools. The instructions and screen-captures in this Section will guide you through installing and manually configuring the software to be used for digital soil mapping procedures for Microsoft Windows desktop platform. Instructions for the other platforms (e.g. Linux Flavours, MacOS) can be found through free online resources.

## Use of R, RStudio and R Packages

**R** is a language and environment for statistical computing. It provides a wide variety of statistical (e.g. linear modeling, statistical tests, time-series, classification, clustering, etc.) and graphical methods, and is highly extensible. 

### Obtaining and Installing R

* Go to https://cloud.r-project.org/index.html to download and install **R**.
* Pick an installation file for your platform. 

### Obtaining and Installing RStudio

Beginners will find it very hard to start using **R** because it has no Graphical User Interface (GUI). There are some GUIs which offer some of the functionality of **R**. **RStudio** makes **R** easier to use. It includes a code editor, debugging and visualization tools, therefore, in this cookbook we would like to focus on the latter including a GUI. You can follow very similar steps to install **RStudio**.

* Go to https://www.rstudio.com/products/rstudio/download/ to download and install **RStudio**'s open source edition.
* On the download page, *RStudio Desktop, Open Source License* option should be selected.
* Pick an installation file for your platform. 

### Getting Started With R	
* **R** Manuals:                  http://cran.r-project.org/manuals.html
* Contributed Documentation:  http://cran.r-project.org/other-docs.html
* Quick-**R**:                    http://www.statmethods.net/index.html
* Stackoverflow **R** Community:  https://stackoverflow.com/questions/tagged/r 

## R Packages

When you download **R**, you get the basic **R** system which implements the **R** language. **R** becomes more useful with the large collection of packages that extend the basic functionality of it. **R** packages are developed by the **R** community.

### Finding R Packages

The primary source for **R** packages is [*CRAN*’s](https://cran.r-project.org/) official website, where currently about 12.000 available packages are listed. For spatial applications, various packages are available. You can obtain information about the available packages directly on *CRAN* with the `available.packages()` function. The function returns a matrix of details corresponding to packages currently available at one or more repositories. An easier way to browse the list of packages is using the [*Task Views*](https://cran.r-project.org/web/views/) link, which groups together packages related to a given topic. 

### Most Used R Packages for Digital Soil Mapping 

As was previously mentioned, **R** is extensible through packages. **R** packages are collections of **R** functions, data, documentation and compiled code easy to share with others. In the following Subsections, we are going to present the most used packages related to digital soil property mapping.

#### Soil Science and Pedometrics{#SoilPedometrics}

[**aqp**](https://CRAN.R-project.org/package=aqp): Algorithms for quantitative pedology.  A collection of algorithms related to modeling of soil resources, soil classification, soil profile aggregation, and visualization.

[**GSIF**](https://CRAN.R-project.org/package=GSIF): Global Soil Information Facility (GSIF). Tools, functions and sample datasets for digital soil mapping. GSIF tools (standards and functions) and sample datasets for global soil mapping.

[**ithir**](http://ithir.r-forge.r-project.org/): A collection of functions and algorithms specific to pedometrics. The package was developed by Brendan Malone at the University of Sydney.

[**soiltexture**](https://CRAN.R-project.org/package=soiltexture): The [*Soil Texture Wizard*](https://cran.r-project.org/web/packages/soiltexture/vignettes/soiltexture_vignette.pdf) is a set of **R** functions designed to produce texture triangles (also called texture plots, texture diagrams, texture ternary plots), classify and transform soil textures data. These functions virtually allow to plot any soil texture triangle (classification) into any triangle geometry (isosceles, right-angled triangles, etc.). The set of functions is expected to be useful to people using soil textures data from different soil texture classification or different particle size systems. Many (> 15) texture triangles from all around the world are predefined in the package. A simple text-based GUI is provided: `soiltexture_gui()`.

#### Spatial Analysis

[**sp**](https://CRAN.R-project.org/package=sp): The package provides classes and methods for spatial data. The classes document where the spatial location information resides, for 2D or 3D data.

[**raster**](https://CRAN.R-project.org/package=raster): Reading, writing, manipulating, analyzing and modeling of gridded spatial data. The package implements basic and high-level functions, processing of very large files is supported.

[**rgdal**](https://CRAN.R-project.org/package=rgdal): Provides bindings to Frank Warmerdam’s Geospatial Data Abstraction Library (GDAL).

[**RSAGA**](https://CRAN.R-project.org/package=RSAGA): The package provides access to geocomputing and terrain analysis functions of [*SAGA GIS*](\url{http://www.saga-gis.org/en/index.html} ) from within **R** by running the command line version of *SAGA*.

#### Modeling

[**caret**](https://CRAN.R-project.org/package=caret): Extensive range of functions for training and plotting classification and regression models.

[**Cubist**](https://CRAN.R-project.org/package=Cubist): Regression modeling using rules with added instance-based corrections. Cubist models were developed by Ross Quinlan.

[**C5.0**](https://CRAN.R-project.org/package=C5.0): C5.0 decision trees and rule-based models for pattern recognition. Another model structure developed by Ross Quinlan.

[**gam**](https://CRAN.R-project.org/package=gam): Functions for fitting and working with generalized additive models.

[**nnet**](https://CRAN.R-project.org/package=nnet): Software for feed-forward neural networks with a single hidden layer, and for multinomial log-linear models.

[**gstat**](https://CRAN.R-project.org/package=gstat): Variogram modeling with simple, ordinary and universal point or block (co)kriging, sequential Gaussian or indicator (co)simulation. The package includes variogram and variogram map plotting utility functions.

[**automap**](https://CRAN.R-project.org/package=automap): This package performs an automatic interpolation by automatically estimating the variogram and then calling [**gstat**](https://CRAN.R-project.org/package=gstat).


#### Mapping and Plotting

Both **raster** and **sp** have handy functions for plotting spatial data. Besides using the base plotting functionality, another useful plotting package is [**ggplot2**](https://cran.r-project.org/web/packages/ggplot2/index.html).

[**plotKML**](https://CRAN.R-project.org/package=plotKML): Writes sp-class, spacetime-class, raster-class and similar spatial and spatiotemporal objects to KML following some basic cartographic rules.

[**leaflet**](https://CRAN.R-project.org/package=leaflet): Create and customize interactive maps using the Leaflet JavaScript library and the [**htmlwidgets**](https://cran.r-project.org/web/packages/htmlwidgets/index.html) package. These maps can be used directly from the **R** console, from **RStudio**, in [**Shiny**](https://shiny.rstudio.com/) apps and **RMarkdown** documents.


### Packages Used in This Cookbook
The authors of this cookbook used a number of different **R** packages. All required packages used in the cookbook can be installed using the following code and the `install.packages()` function when starting a new SOC mapping project. Alternatively, the code for the installation of the needed packages is included at the beginning of each Chapter. 

```{r, eval=FALSE}
# Install all required R packages used in the cookbook
install.packages(c("aqp", "automap", "car", "caret", "e1071",
                   "GSIF", "htmlwidgets", "leaflet", "mapview",
                   "Metrics", "openair", "plotKML", "psych", 
                   "quantregForest", "randomForest", "raster",
                   "rasterVis", "reshape", "rgdal", "RSQLite",
                   "snow", "soiltexture", "sf", "sp"))
```


## R and Spatial Data

**R** has a large and growing number of spatial data packages. We recommend taking a quick browse on **R**’s official website to see the spatial packages available: \url{http://cran.r-project.org/web/views/Spatial.html}
 
### Reading Shapefiles 

*ESRI*'s shapefile format is widely used for storing vector-based spatial data (i.e., points, lines, polygons). This example demonstrates use of raster package that provides functions for reading and/or writing shapefiles.

```{r}
library(raster)
# Load the soil map from a shapefile file
soilmap <- shapefile("MK_soilmap_simple.shp")
```


We may want to use these data in other GIS environments such as *ArcGIS*, *QGIS*, *SAGA GIS*, etc. This means we need to export the `SpatialPointsDataFrame` to an appropriate spatial data format such as a shapefile. 

```{r}
# For example, we can select the soil units classified as 
# fluvisols according to WRB
Fluvisols <- soilmap[soilmap$WRB == "Fluvisol",]

# Save this as a new shapefile 
shapefile(Fluvisols, filename = 'results/fluvisols.shp',
          overwrite = TRUE)
```


### Coordinate Reference Systems (CRS) in R

We need to define the CRS (Coordinate Reference System) to be able to perform any sort of spatial analysis in **R**. To clearly tell **R** this information we define the CRS which describes a reference system in a way understood by the PROJ.4 projection library http://trac.osgeo.org/proj/. 

An interface to the PROJ.4 library is available in the **rgdal** package. An alternative to using Proj4 character strings, we can use the corresponding yet simpler EPSG (European Petroleum Survey Group) code. **rgdal** also recognizes these codes. If you are unsure of the Proj4 or EPSG code for the spatial data that you have but know the CRS, you should consult http://spatialreference.org/ for assistance.

```{r}
# Print the CRS for the object soilmap
soilmap@proj4string
```

The following example shows how you can create a spatial object from a .csv file. We can use the `coordinates()` function from the **sp** package to define which columns in the data frame refer to actual spatial coordinates—here the coordinates are listed in columns X and Y.

```{r}
# Load the table with the soil observations site information
dat_sites <- read.csv(file = "data/site-level.csv")

# Convert from table to spatial points object
coordinates(dat_sites) <- ~ X + Y

# Check the coordinate system
dat_sites@proj4string

# As the CRS is not defined, we can assign the correct CRS as we
# have information about it. In this case, it should be EPSG:4326
dat_sites@proj4string <- CRS("+init=epsg:4326")

# Check the CRS again
dat_sites@proj4string
```


### Working with Rasters

Most of the functions for handling raster data are available in the **raster** package. There are functions for reading and writing raster files from and to different formats. In digital soil mapping, we mostly work with data in table format and then rasterize this data so that we can produce a continuous map. For doing this in **R** environment, we will load raster data in a data frame. This data is a digital elevation model (DEM) provided by [*ISRIC*](http://www.isric.org/) for Former Yugoslav Republic of Macedonia (FYROM).

```{r}
# For handling raster data, we load raster package
library(raster)

# Load DEM from tif file
DEM <- raster("covs/DEMENV5.tif")
```


We may want to export this raster to a suitable format to work in a standard GIS environment. See the help file for writing a raster `?writeRaster` to get information regarding the supported grid types that data can be exported into. Here, we will export our raster to *ESRI* Ascii, as it is a common and universal raster format.

We may also want to export our mac.dem to KML file using the `KML()` function. `KML()` is a handy function from the **raster** package for exporting grids to kml format. Note that we need to re-project the data to WGS84 geographic. The raster re-projection is performed using the `projectRaster()` function. Look at the help file `?projectRaster` for this.

## Other DSM-Related Software and Tools

* *QGIS*:      Available at http://www.qgis.org/en/site/forusers/download.html 
* *SAGA GIS*:  Available at https://sourceforge.net/projects/saga-gis/files/
